---
title: An introduction to R
subtitle: AITI Talk
format:
  ubd40-revealjs:
    # self-contained: false
    # chalkboard: true
    # toc: true
    # toc-depth: 1
    transition: fade
    auto-stretch: false
    width: 1250  # 1050
    height: 760  # 700
    self-contained: false
    chalkboard: true
    toc: false
    toc-depth: 1
    multiplex: true
    code-block-height: 700px
    # html-table-processing: none
author:
  - name: Haziq Jamil
    orcid: 0000-0003-3298-1010
    email: haziq.jamil@ubd.edu.bn
    affiliations: Universiti Brunei Darussalam
date: 2025-05-21
# bibliography: refs.bib
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: setup
library(tidyverse)  # data wrangling tools
library(tinyplot)
library(tidytext)   # bigrams
library(tm)         # text mining
library(wordcloud)  # word clouds
tinytheme("clean2")
```


## Introduction

## What is R?

What can we do using R?
Mainly used for statistical analysis.
Focus on data wrangling and data analysis.
Visualisations.
Create reproducible reports.

## General purpose programming language

Show for loops?


## There's a package for that


`{ellmer}`
`{rvest}`
`{bruneimap}`


## Getting started

- RStudio vs R [IDE vs language]
- Quick lay of the land
- RProj and "home" directory
- Mindset shift: Code everything!!! Reproducibility matters

## Importing data



```{r}
dat <- read_csv("fake_survey.csv")
glimpse(dat)
```

## Data types

- Numeric (continuous vs integer)
- Logical
- Character
- Factors
- Ordered

## Summary statistics

### Continuous data 

```{r}
x <- dat$age
head(x)

# Mean and standard deviation
mean(x)
sd(x)

# Quick summary of the data
summary(x)
```

## Histograms

```{r}
#| out-width: 100%
#| fig-height: 5.2
tinyplot(x, type = "histogram", main = "Histogram of Age", xlab = "Age", 
         ylab = "Frequency")
```

## Density plots

```{r}
#| out-width: 100%
#| fig-height: 5.2
tinyplot(~ age | gender, data = dat, type = "density", main = "Age distribution", 
         xlab = "Age", fill = "by")
```



